Using simulation to optimize the sampling design of fisheries-independent trawl surveys

### 2018-09-20 analysis/paper/sim_survey_paper.Rmd
Add two new figures to the paper given new length-based output

```{r data, include = FALSE}
library(SimSurvey)
library(plotly)
library(raster)
library(lattice)
library(viridis)
library(data.table)
load("analysis/cod_sim_exports/2018-09-14_test/test_output.RData")
```

@@

## little helper function for the fan plot
fan_plot <- function(d, x = "Age", total_strat = FALSE) {

  if (total_strat) {
    ints_by = c("year", "set_den")
  } else {
    ints_by = c("year", "age", "set_den", "lengths_cap")
  }

@@

  if (x == "Age") {
    p <- ggplot(data = ints, aes(x = age))
    xlabs <- c("2", "4", "6", "8+")
    xbreaks <- seq(2, 8 , 2)
  }
  if (x == "Year") {
    p <- ggplot(data = ints, aes(x = year))
    xlabs <- waiver()
    xbreaks <- waiver()
  }

@@

  if (total_strat) {
    p <- p + facet_wrap(~ set_den, nrow = 1,
                        labeller = label_bquote(cols = D[sets]==.(set_den)))
  } else {
    p <- p + facet_grid(lengths_cap ~ set_den, 
                        labeller = label_bquote(cols = D[sets]==.(set_den),
                                                rows = M[lengths]==.(lengths_cap)))
  }

@@

$$N_{a,y} = N_{a-1,y-1} e^{-Z_{a-1,y-1}}$$
Here, numbers-at-age in the first year are filled via exponential decay, $N_{a,1} = N_{a-1,1} e^{-Z_{a-1,1}}$, numbers at age 1 (i.e. recruits) vary around a baseline value, $log(N_{1,y}) = log(\mu_r) + \epsilon_{y}$, and total mortality is set to a baseline level plus process error, $log(Z_{a,y}) = log(\mu_Z) + \delta_{a,y}$. The error around the recruitment process was set to follow a random walk, $\epsilon_{y} \sim N(\epsilon_{y - 1}, \sigma_{r}^{2})$, and the process error was simulated using the covariance structure described in Cadigan [-@Cadigan2016], $\delta_{a,y} \sim N(0, \Sigma_{a,y})$. The covariance across ages and years is controlled by a process error variance parameter ($\sigma_{\delta}^2$) along with age and year correlation parameters ($\varphi_{\delta, age}$ and $\varphi_{\delta, year}$, respectively). This structure allows for autocorrelation in process errors across ages and years (i.e. total mortality can be made to be more similar for fish that are closer together in age and time). Abundance-at-age is then converted to abundance-at-length using the original von Bertalanffy growth curve [@Cailliet2006]:

@@

Where $l_{\infty}$ is the mean asymptotic length, $l_0$ is length at birth, $K$ is the growth rate parameter and the error is assumed to follow the normal distribution, $\varepsilon \sim N(0, \sigma_{l}^2)$. Abundance in discrete length groups, $l_{group}^{N}$, was determined by using this formula to calculate the probability of being within a specific length group given age and then using the resultant length-age-key to convert abundance-at-age to abundance-at-length. Overall, this  formulation allows for the simulation of an array of population dynamics that one might expect to observe.

@@

The next step in the simulation is to distribute the abundance-at-age matrix simulated using the cohort model throughout a spatial field. Here, a grid of $s$ locations was generated with a resolution of $R$ and a depth gradient. The depth gradient was set-up to have a sigmoid shape with a depth range of $[d_{min}, d_{max}]$, shelf depth of $d_{shelf}$ and a shelf width of $w_{shelf}$. The simulated population is distributed through the grid by simulating spatial-temporal noise controlled by a parabolic relationship with depth and covariance between ages, years and space

@@

Ideally the parameter settings used here would be based on model fits to real data, however, conditioning the operating model described above was outside the scope of this project. Moreover, many of the parameters described above may not be identifiable or it may be difficult to obtain un-biased estimates [e.g. bias introduced by length-stratified sampling complicate the fitting of growth models; @Morgan1997]. As such, parameters were informally selected to roughly correspond to observations typically obtained by the Fisheries and Oceans Canada survey of cod in NAFO division 3Ps, Newfoundland. Though this approach is far from exact, the goal was to approximate the life-history and distribution of an actual population to simulate an arena for testing the consequences of surveys with different sampling protocol. Parameter settings and descriptions used in this case study are presented in Table \@ref(tab:par). A series of `r nrow(sim$surveys)` surveys were tested using an array of different sampling protocol outlined in Table \@ref(tab:tests). Each survey was simulated `r max(sim$age_strat_error$sim)` times over the same population and samples were run through a stratified analysis [@Smith1981] to obtain annual estimates of total abundance ($\hat{I}_{y,k}$), abundance-at-length ($\hat{I}_{l,y,k}$) and abundance-at-age ($\hat{I}_{a,y,k}$). We used root-mean-squared error (RMSE) as a measure of the precision and bias of the abundance-at-age estimates from each survey:

$$RMSE = \sqrt{\frac{ \sum_{a = 1}^{N_a} \sum_{y = 1}^{N_y} \sum_{k = 1}^{N_k} (\hat{I}_{a,y,k} - I_{a,y})^2}{N_a N_y N_k}}$$
Where $N_a$, $N_y$, and $N_k$ are the number of years, ages and simulations, respectively, and $I_{a,y}$ is the true abundance available to the survey (i.e. catchability corrected abundance; $I_{a,y} = q_a N_{a,y}$). Intra-haul correlation (i.e. intraclass correlation) of set catch was also calculated by fitting a mixed-model to the lengths data with set as a random effect [fit using the lme4 package; @Bates2015]; variance estimates from this model were then plugged into the ICC(1) formula presented in McGraw and Wong [-@Mcgraw1996]. All simulations and analyses were conducted using `r substr(version$version.string, 1, 15)` [@R2017] and the code has been developed into a package called `SimSurvey` to facilitate future work and collaboration (https://github.com/PaulRegular/SimSurvey).

@@

- Feedback and advice: Alejandro Buren, Noel Cadigan, Karen Dwyer, Geoff Evans, Brian Healey, Paul Higdon, Danny Ings, Mariano Koen-Alonso, Joanne Morgan, Derek Osborne, Dwayne Pittman, Don Power, Martha Robertson, Mark Simpson, Brad Squires, Don Stansbury, Peter Upward
- Dave Cote for a review of a previous version of this manuscript

@@

p <- fan_plot(d, x = "Year", total_strat = TRUE)
cowplot::save_plot(plot = p, filename = "analysis/paper/figures/total_strat_true_vs_est.pdf",
                   base_height = 3, base_width = 8)
p

@@

@@

@@

```{r error, cache = TRUE, fig.height = 3, fig.width = 8.5, results = "hide", fig.cap = "- Wireframe plot of root-mean-squared error (RMSE) from an array of surveys with different sampling protocol. Panels represent surveys with different set densities ($D_{sets}$), x-axes represent the maximum sampling effort of lengths per set ($M_{lengths}$), and y-axes represent the maximum number of ages to collect per length group ($M_{ages}$)."}

@@

```{r appen-data}
## this export holds the same distribution test
load("analysis/cod_sim_exports/2018-09-07_test/test_output.RData")
```

>>>>

```{r data, include = FALSE}
library(SimSurvey)
library(plotly)
library(raster)
library(lattice)
library(viridis)
library(data.table)
load("analysis/cod_sim_exports/2018-09-17_age_clust_test/test_output.RData")
```

@@

## little helper function for the fan plot
fan_plot <- function(d, x = "Age", which_strat = "Age") {

  if (which_strat == "Total") {
    ints_by = c("year", "set_den")
  }
  if (which_strat == "Age") {
    ints_by = c("year", "age", "set_den", "lengths_cap")
  }
  if (which_strat == "Length") {
    ints_by = c("year", "length", "set_den", "lengths_cap")
  }

@@

  if (x == "Age") {
    p <- ggplot(data = ints, aes(x = age))
    xlabs <- c("2", "4", "6", "8", "10+")
    xbreaks <- seq(2, 10, 2)
  }
  if (x == "Length") {
    p <- ggplot(data = ints, aes(x = length))
    xbreaks <- xlabs <- seq(0, 100, 20)
    xlabs[xlabs == 100] <- "100+"
  }
  if (x == "Year") {
    p <- ggplot(data = ints, aes(x = year))
    xlabs <- waiver()
    xbreaks <- waiver()
  }

@@

  if (which_strat == "Total") {
    p <- p + facet_wrap(~ set_den, nrow = 1,
                        labeller = label_bquote(cols = D[sets]==.(set_den)))
  } else {
    p <- p + facet_grid(lengths_cap ~ set_den, 
                        labeller = label_bquote(cols = D[sets]==.(set_den),
                                                rows = M[lengths]==.(lengths_cap)))
  }

@@

$$N_{a,y} = N_{a-1,y-1} e^{-Z_{a-1,y-1}}$$
Here, numbers at age in the first year are filled via exponential decay, $N_{a,1} = N_{a-1,1} e^{-Z_{a-1,1}}$, numbers at age 1 (i.e. recruits) vary around a baseline value, $log(N_{1,y}) = log(\mu_r) + \epsilon_{y}$, and total mortality is set to a baseline level plus process error, $log(Z_{a,y}) = log(\mu_Z) + \delta_{a,y}$. The error around the recruitment process was set to follow a random walk, $\epsilon_{y} \sim N(\epsilon_{y - 1}, \sigma_{r}^{2})$, and the process error was simulated using the covariance structure described in Cadigan [-@Cadigan2016], $\delta_{a,y} \sim N(0, \Sigma_{a,y})$. The covariance across ages and years is controlled by a process error variance parameter ($\sigma_{\delta}^2$) along with age and year correlation parameters ($\varphi_{\delta, age}$ and $\varphi_{\delta, year}$, respectively). This structure allows for autocorrelation in process errors across ages and years (i.e. total mortality can be made to be more similar for fish that are closer together in age and time). Abundance at age is then converted to abundance at length using the original von Bertalanffy growth curve [@Cailliet2006]:

@@

Where $l_{\infty}$ is the mean asymptotic length, $l_0$ is length at birth, $K$ is the growth rate parameter and the error is assumed to follow the normal distribution, $\varepsilon \sim N(0, \sigma_{l}^2)$. Abundance in discrete length groups, $l_{group}^{N}$, was determined by using this formula to calculate the probability of being within a specific length group given age and then using the resultant length-age-key to convert abundance at age to abundance at length. Overall, this  formulation allows for the simulation of an array of population dynamics that one might expect to observe.

@@

The next step in the simulation is to distribute the abundance at age matrix simulated using the cohort model throughout a spatial field. Here, a grid of $s$ locations was generated with a resolution of $R$ and a depth gradient. The depth gradient was set-up to have a sigmoid shape with a depth range of $[d_{min}, d_{max}]$, shelf depth of $d_{shelf}$ and a shelf width of $w_{shelf}$. The simulated population is distributed through the grid by simulating spatial-temporal noise controlled by a parabolic relationship with depth and covariance between ages, years and space

@@

Ideally the parameter settings used here would be based on model fits to real data, however, conditioning the operating model described above was outside the scope of this project. Moreover, many of the parameters described above may not be identifiable or it may be difficult to obtain un-biased estimates [e.g. bias introduced by length-stratified sampling complicate the fitting of growth models; @Morgan1997]. As such, parameters were informally selected to roughly correspond to observations typically obtained by the Fisheries and Oceans Canada survey of cod in NAFO division 3Ps, Newfoundland. Though this approach is far from exact, the goal was to approximate the life-history and distribution of an actual population to simulate an arena for testing the consequences of surveys with different sampling protocol. Parameter settings and descriptions used in this case study are presented in Table \@ref(tab:par). A series of `r nrow(sim$surveys)` surveys were tested using an array of different sampling protocol outlined in Table \@ref(tab:tests). Each survey was simulated `r max(sim$age_strat_error$sim)` times over the same population and samples were run through a stratified analysis [@Smith1981] to obtain annual estimates of total abundance ($\hat{I}_{y,k}$), abundance at length ($\hat{I}_{l,y,k}$) and abundance at age ($\hat{I}_{a,y,k}$). We used root-mean-squared error (RMSE) as a measure of the precision and bias of the abundance at age estimates from each survey:

$$RMSE = \sqrt{\frac{ \sum_{a = 1}^{N_a} \sum_{y = 1}^{N_y} \sum_{k = 1}^{N_k} (\hat{I}_{a,y,k} - I_{a,y})^2}{N_a N_y N_k}}$$
Where $N_a$, $N_y$, and $N_k$ are the number of years, ages and simulations, respectively, and $I_{a,y}$ is the true abundance available to the survey (i.e. catchability corrected abundance; $I_{a,y} = q_a N_{a,y}$). RMSE was also calculated for abundance at length estimates. Intra-haul correlation (i.e. intraclass correlation) of set catch was also calculated by fitting a mixed-model to the lengths data with set as a random effect [fit using the lme4 package; @Bates2015]; variance estimates from this model were then plugged into the ICC(1) formula presented in McGraw and Wong [-@Mcgraw1996]. All simulations and analyses were conducted using `r substr(version$version.string, 1, 15)` [@R2017] and the code has been developed into a package called `SimSurvey` to facilitate future work and collaboration (https://github.com/PaulRegular/SimSurvey).

@@

- Feedback and advice: Aaron Adamack, Alejandro Buren, Noel Cadigan, Karen Dwyer, Geoff Evans, Brian Healey, Paul Higdon, Danny Ings, Mariano Koen-Alonso, Joanne Morgan, Derek Osborne, Dwayne Pittman, Don Power, Martha Robertson, Mark Simpson, Brad Squires, Don Stansbury, Peter Upward
- Dave Cote for a review of a previous version of this manuscript

@@

p <- fan_plot(d, x = "Year", which_strat = "Total")
cowplot::save_plot(plot = p, filename = "analysis/paper/figures/total_strat_true_vs_est.pdf",
                   base_height = 3, base_width = 8)
p

@@

```{r length, cache = TRUE, fig.width = 8, fig.height = 6, fig.cap = "- Fanchart of stratified estimates of abundance at length from year seven of the simulation from surveys with different set densities, $D_{sets}$, and length sampling protocol, $M_{lengths}$. The heavy dark line indicates the true population available to the survey."}

sub_surveys <- sim$surveys[sim$survey$set_den %in% which_set_dens &
                             sim$survey$lengths_cap %in% which_len_caps & 
                             sim$survey$ages_cap == 10, ]
sub_strat <- sim$length_strat_error[eval(sim$length_strat_error$survey %in% sub_surveys$survey &
                                           sim$length_strat_error$year == which_year), ]
d <- merge(sub_surveys, sub_strat, by = "survey")
d$length[d$length > 100] <- 100 # aggregate beyond 100 cm
d <- d[, list(I_hat = sum(I_hat), I = sum(I)), 
        by = c("year", "length", "sim", "survey", "set_den", "lengths_cap")]

p <- fan_plot(d, x = "Length", which_strat = "Length")
cowplot::save_plot(plot = p, filename = "analysis/paper/figures/true_vs_est_by_length.pdf",
                   base_height = 6, base_width = 8)
p

@@

sub_surveys <- sim$surveys[sim$survey$set_den %in% which_set_dens &
                             sim$survey$lengths_cap %in% which_len_caps & 
                             sim$survey$ages_cap == 10, ]
sub_strat <- sim$age_strat_error[eval(sim$age_strat_error$survey %in% sub_surveys$survey &
                                        sim$age_strat_error$year == which_year), ]
d <- merge(sub_surveys, sub_strat, by = "survey")
d <- d[d$year == which_year & d$set_den %in% which_set_dens &
         d$lengths_cap %in% which_len_caps & d$ages_cap == 10, ]
d$age[d$age > 10] <- 10 # aggregate beyond 10
d <- d[, list(I_hat = sum(I_hat), I = sum(I)), 
       by = c("year", "age", "sim", "survey", "set_den", "lengths_cap")]

@@

```{r error-lines, cache = TRUE, fig.height = 3.8, fig.width = 8.5, results = "hide", fig.cap = "- Root-mean-squared error (RMSE) of abundance at length and abundance at age estimates as a function of the number of fish measured ($N_{lengths}$) from an array of surveys with different sampling protocol. The labels represent surveys with different set densities ($D_{sets}$), where number of sets are shown in brackets, and the lines are color coded by length sampling protocol ($M_{lengths}$). Number of ages sampled per length group, $M_{ages}$, was 10 in all scenarios."}

totals <- sim$samp_totals[, list(n_sets = mean(n_sets), n_caught = mean(n_caught),
                                 n_measured = mean(n_measured), n_aged = mean(n_aged)),
                          by = "survey"]

errors <- merge(sim$surveys, sim$length_strat_error_stats, by = "survey")
length_d <- merge(errors, totals, by = "survey")
length_d <- length_d[length_d$ages_cap == 10, ]
length_d$strat <- "Abundance at length"

errors <- merge(sim$surveys, sim$age_strat_error_stats, by = "survey")
age_d <- merge(errors, totals, by = "survey")
age_d <- age_d[age_d$ages_cap == 10, ]
age_d$strat <- "Abundance at age"

d <- rbind(length_d, age_d)
d$cap <- factor(d$lengths_cap)
d$strat <- factor(d$strat, levels = c("Abundance at length", "Abundance at age"))

lab_d <- d[lengths_cap == max(lengths_cap),
           list(lengths_cap, n_measured, n_sets, cap, RMSE),
           by = c("set_den", "strat")]
lab_d$lab <- paste0(formatC(lab_d$set_den), " (", lab_d$n_sets, ")")

p <- ggplot(data = d, aes(x = n_measured, y = RMSE, group = set_den)) +
  geom_line(aes(color = cap), size = 2, lineend = "round") +
  geom_text(data = lab_d, aes(label = lab), nudge_x = 500, hjust = 0) +
  geom_point(data = lab_d, aes(color = cap)) +
  xlab(bquote(N[lengths])) +
  scale_y_continuous(labels = format_si()) +
  scale_x_continuous(labels = format_si(), limits = c(0, max(d$n_measured) * 1.3),
                     breaks = pretty(d$n_measured, 6)) +
  scale_color_viridis(name = bquote(M[lengths]), discrete = TRUE) +
  facet_wrap(~ strat, scales = "free", dir = "h") +
  theme_bw() +
  theme(legend.justification = c(0, 1),
        plot.margin = unit(c(0.2, 0.2, 0.2, 0.2), "cm"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_blank(),
        panel.border = element_rect(fill = NA, color = "darkgrey", size = 0.4),
        axis.ticks = element_line(color = "darkgrey", size = 0.4))

cowplot::save_plot(plot = p, filename = "analysis/paper/figures/error_lines.pdf",
                   base_height = 3.8, base_width = 8.5)
p

```{r error-wireframe, cache = TRUE, fig.height = 3, fig.width = 8.5, results = "hide", fig.cap = "- Wireframe plot of root-mean-squared error (RMSE) from an array of surveys with different sampling protocol. Panels represent surveys with different set densities ($D_{sets}$), x-axes represent the maximum sampling effort of lengths per set ($M_{lengths}$), and y-axes represent the maximum number of ages to collect per length group ($M_{ages}$)."}

@@

```{r appen-data}
## this export holds the same distribution test
load("analysis/cod_sim_exports/2018-09-18_no_age_clust_test/test_output.RData")
```
